Accelerated Bilateral Filtering for Multi-core Architectures and Code Unified Device Architecture

### 2012-03-11 LateX/paper8.tex
changed manycore to many-core. changing title to be more specific.

\begin{document}
\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:

@@

In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory. Each warp executes one instruction at a time for its own group of threads. Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls. Hence, synchronization between threads in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that meet the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*j_pitch + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.

\subsection{Shared Memory and Tile Processing}
\begin{figure}
\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
\caption{An illustrated overview of the bilateral filtering process. The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
\label{fig:pspro}
\end{figure}

@@

On the GPU, our pair-symmetric algorithm outperformed our naive algorithm for every set of tests. For example, Figure~\ref{fig:hw10} shows the pair-symmetric algorithm's best timing to be roughly 75\% of the naive algorithm's best timing for processing a 3000 x 3000 image using a 21 x 21 pixel filter window. The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size. In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm. Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased. Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.

While our tests returned a maximum 25\% reduction in execution time, our pair-symmetric algorithm has the potential to reach a near 50\% reduction in execution time. Furthermore, the benchmarks of our pair-symmetric algorithm reveal minor irregularities in the trajectory of performance derived from execution timings versus respective incremental adjustments of parameter values such as CUDA block and grid dimensions. The main cause of both of these observations is shared memory. To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time. Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings. To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

Another challenge in interpreting our results was porting the concept of "speed-up" to general purpose GPU programming. Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used---that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper---and we also made another reference timing by running our GPU implementations using a single thread. For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms. With respect to these reference timings, the speed-up is roughly 20,500x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.

>>>>

\begin{document}
\title{Accelerated Bilateral Filtering for Multi-core Architectures and Code Unified Device Architecture}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and many-core acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel. For many-core architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel. A formal expression is given below:

@@

In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory. Each warp executes one instruction at a time for its own group of threads. Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls. Therefore, synchronization between threads in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.

\subsection{Shared Memory and Tile Processing}
\begin{figure}
\includegraphics[width = 1\columnwidth]{images/project23_.pdf}
\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 1\columnwidth]{images/tiles19.pdf}
\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 1\columnwidth]{images/smallfaces5.pdf}
\caption{An illustrated overview of the bilateral filtering process. The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
\label{fig:pspro}
\end{figure}

@@

On the GPU, our pair-symmetric algorithm outperformed our naive algorithm for every set of tests. For example, Figure~\ref{fig:hw10} shows the pair-symmetric algorithm's best timing to be roughly 75\% of the naive algorithm's best timing for processing a 3000 x 3000 image using a 21 x 21 pixel filter window. The test that gave the closest difference in timing as shown in Figure~\ref{fig:hw10light} was processing a 768 x 712 pixel image using a 11 x 11 window size. In this test, the pair-symmetric algorithm's best timing was at a virtual tie with the naive algorithm's best timing at 0.00390 seconds, but upon further testing we found the pair-symmetric algorithm's optimal timing to be 0.00388 seconds when changing the grid dimension to 16 x 16 concurrent blocks, thus slightly outperforming the naive algorithm. Figures \ref{fig:hw10} and \ref{fig:hw10light} show exponential rises in execution time for both algorithms as block vertical dimension decreased. Due to runtime complexity of the bilateral filter being the square of the filter width, Figures (b) and (d), which reflect our tests that used a 5 pixel filter width, do not exhibit the steep curvature that Figures (a) and (c) exhibit.

While our tests returned a maximum 25\% reduction in execution time, our pair-symmetric algorithm has the potential to reach a near 50\% reduction in execution time. Furthermore, the benchmarks of our pair-symmetric algorithm reveal minor irregularities in the trajectory of performance derived from execution timings versus respective incremental adjustments of parameter values such as CUDA block and grid dimensions. The main cause of both of these observations is shared memory. To better understand the impact of the additional shared memory transactions that our pair-symmetric algorithm requires, whereas the naive algorithm does not require shared memory for output storage, we implemented a single-threaded CPU-based version of both the naive bilateral filter algorithm and pair-symmetric algorithm so that native-memory (RAM, L1, or L2 cache) could substitute for the GPU's shared memory. The results were remarkably illustrative of our pair-symmetric algorithm's potential. The naive algorithm running on the CPU with a filter width of 10 pixels processed a 3000 x 3000 image in an average of 1206.7 seconds while our pair-symmetric algorithm took 611.8 seconds - a 49.23\% reduction in execution time. Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., reducing timings by no more than than 25\%. To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code, CUDA's quasi-assembly language, of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

Another challenge in interpreting our results was porting the concept of "speed-up" to general purpose GPU programming. Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used---that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper---and we also made another reference timing by running our GPU implementations using a single thread. For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms. With respect to these reference timings, the speed-up is roughly 21,090x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.
